The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-08-14 talk/vmil2012/paper.tex
replace front- and backend with high- and low-level components

%___________________________________________________________________________
\todo{find a better name for \texttt{low-level resume data}}
\todo{find better names for JIT front- and backend}
\todo{mention somewhere that it is to be expected that most guards do not fail}
\section{Introduction}

@@

The set of central concepts upon which this work is based is described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe for RPython's tracing JIT the details of guards in
the frontend\bivab{better term for this?} related to recording and storing the
information required to rebuild the interpreter state in case of a guard
failure. Once the frontend has traced and optimized a loop it invokes the
backend to compile the operations to machine code, Section \ref{sec:Guards in
the Backend} describes the low-level aspects of how guards are implemented in
the JIT-backend. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

\section{Guards in the Frontend} %{Resume Data}
\label{sec:Resume Data}

@@

\begin{figure}
    \input{figures/log.tex}
    \caption{Optimized trace}
    \label{fig:trace-log}
\end{figure}
% section Resume Data (end)
\section{Guards in the Backend}
\label{sec:Guards in the Backend}

@@

After optimization the resulting trace is handed over to the platform specific
backend to be compiled to machine code. The compilation phase consists of two
passes over the lists of instructions, a backwards pass to calculate live
ranges of IR-level variables and a forward pass to emit the instructions. During
the forward pass IR-level variables are assigned to registers and stack
locations by the register allocator according to the requirements of the
emitted instructions.  Eviction/spilling is performed based on the live range
information collected in the first pass. Each IR instruction is transformed
into one or more machine level instructions that implement the required
semantics. Operations without side effects whose result is not used are not
emitted. Guards instructions are transformed into fast checks at the machine
code level that verify the corresponding condition.  In cases the value being
checked by the guard is not used anywhere else the guard and the operation
producing the value can often be merged, further reducing the overhead of the guard.
Figure \ref{fig:trace-compiled} shows how the \texttt{int\_eq} operation
followed by a \texttt{guard\_false} from the trace in Figure~\ref{fig:trace-log} are compiled to
pseudo-assembler if the operation and the guard are compiled separated or if
they are merged.

@@

Using the encoded location information the bailout handler reads from the
saved execution state the values that the IR-variables had  at the time of the
guard failure and stores them in a location that can be read by the frontend.
After saving the information the control is passed to the frontend signaling
which guard failed so the frontend can read the information passed and restore
the state corresponding to the point in the program.

As in previous sections the underlying idea for the design of guards is to have
a fast on-trace profile and a potentially slow one in the bailout case where
the execution has to return to the interpreter due to a guard failure. At the same
time the data stored in the backend, required to rebuild the state, should be as
compact as possible to reduce the memory overhead produced by the large number
of guards, the numbers in Figure~\ref{fig:backend_data} illustrate that the
compressed encoding currently has about 15\% to 25\% of the size of of the
generated instructions on x86.

@@

The results presented in this section are based on numbers gathered by running
a subset of the standard PyPy benchmarks. The PyPy benchmarks are used to
measure the performance of PyPy and are composed of a series of
micro-benchmarks and larger programs.\footnote{\url{http://speed.pypy.org/}} The
benchmarks were taken from the PyPy benchmarks repository using revision
\texttt{ff7b35837d0f}.\footnote{\url{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}}
The benchmarks were run on a version of PyPy based on the
tag~\texttt{0b77afaafdd0} and patched to collect additional data about the
guards in the machine code
backends.\footnote{\url{https://bitbucket.org/pypy/pypy/src/0b77afaafdd0}} The
tools used to run and evaluate the benchmarks including the patches applied to
the PyPy sourcecode can be found in the repository for this
paper.\footnote{\url{https://bitbucket.org/pypy/extradoc/src/tip/talk/vmil2012}}
All
benchmark data was collected on a MacBook Pro 64 bit running Max OS 10.8 with
the loop unrolling optimization disabled.\footnote{Since loop unrolling
duplicates the body of loops it would no longer be possible to meaningfully
compare the number of operations before and after optimization. Loop unrolling
is most effective for numeric kernels, so the benchmarks presented here are not
affected much by its absence.}

@@

The overhead that is incurred by the JIT to manage the \texttt{resume data},
the \texttt{low-level resume data} as well as the generated machine code is
shown in Figure~\ref{fig:backend_data}. It shows the total memory consumption
of the code and of the data generated by the machine code backend and an
approximation of the size of the \texttt{resume data} structures for the
different benchmarks mentioned above. The machine code taken into account is
composed of the compiled operations, the trampolines generated for the guards
and a set of support functions that are generated when the JIT starts and which
are shared by all compiled traces. The size of the \texttt{low-level resume
data} is the size of the compressed mapping from registers and stack to
IR-level variables and finally the size of the \texttt{resume data} is an
approximation of the size of the compressed high-level resume data as described
in Section~\ref{sec:Resume Data}.\footnote{
The size of the resume data is not measured at runtime, but reconstructed from
log files.}

>>>>

%___________________________________________________________________________
\todo{find a better name for \texttt{low-level resume data}}
\todo{mention somewhere that it is to be expected that most guards do not fail}
\section{Introduction}

@@

The set of central concepts upon which this work is based is described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe for RPython's tracing JIT the details of guards in
the high-level, target independant, component of the JIT related to recording and storing the
information required to rebuild the interpreter state in case of a guard
failure. Once the high-level component has traced and optimized a loop it invokes the
low-level component to compile the operations to machine code, Section \ref{sec:Guards in
the Backend} describes the aspects of the implementaion of guards in
the low-level JIT component. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

\section{Guards in the High-Level JIT Components} %{Resume Data}
\label{sec:Resume Data}

@@

\begin{figure}
    \input{figures/log.tex}
    \caption{Optimized trace}
    \label{fig:trace-log}
\end{figure}
% section Resume Data (end)
\section{Guards in the Low-Level JIT Component}
\label{sec:Guards in the Backend}

@@

After optimization the resulting trace is handed over to the platform specific
low-level implementation to be compiled to machine code. The compilation phase consists of two
passes over the lists of instructions, a backwards pass to calculate live
ranges of IR-level variables and a forward pass to emit the instructions. During
the forward pass IR-level variables are assigned to registers and stack
locations by the register allocator according to the requirements of the
emitted instructions.  Eviction/spilling is performed based on the live range
information collected in the first pass. Each IR instruction is transformed
into one or more machine level instructions that implement the required
semantics. Operations without side effects whose result is not used are not
emitted. Guards instructions are transformed into fast checks at the machine
code level that verify the corresponding condition.  In cases the value being
checked by the guard is not used anywhere else the guard and the operation
producing the value can often be merged, further reducing the overhead of the guard.
Figure \ref{fig:trace-compiled} shows how the \texttt{int\_eq} operation
followed by a \texttt{guard\_false} from the trace in Figure~\ref{fig:trace-log} are compiled to
pseudo-assembler if the operation and the guard are compiled separated or if
they are merged.

@@

Using the encoded location information the bailout handler reads from the
saved execution state the values that the IR-variables had at the time of the
guard failure and stores them in a location that can be read by the high-level component.
After saving the information the control is returned to the high-level component signaling
which guard failed. The stored low-level information can be read and used to restore
the execution state corresponding to the guard failure in the interpreter.

As in previous sections the underlying idea for the design of guards is to have
a fast on-trace profile and a potentially slow one in the bailout case where
the execution has to return to the interpreter due to a guard failure. At the same
time the data stored in the low-level component, required to rebuild the state, should be as
compact as possible to reduce the memory overhead produced by the large number
of guards, the numbers in Figure~\ref{fig:backend_data} illustrate that the
compressed encoding currently has about 15\% to 25\% of the size of of the
generated instructions on x86.

@@

The results presented in this section are based on numbers gathered by running
a subset of the standard PyPy benchmarks. The PyPy benchmarks are used to
measure the performance of PyPy and are composed of a series of
micro-benchmarks and larger programs.\footnote{\url{http://speed.pypy.org/}} The
benchmarks were taken from the PyPy benchmarks repository using revision
\texttt{ff7b35837d0f}.\footnote{\url{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}}
The benchmarks were run on a version of PyPy based on
revision~\texttt{0b77afaafdd0} and patched to collect additional data about the
guards in the generated machine code.\footnote{\url{https://bitbucket.org/pypy/pypy/src/0b77afaafdd0}} The
tools used to run and evaluate the benchmarks including the patches applied to
the PyPy sourcecode can be found in the repository for this
paper.\footnote{\url{https://bitbucket.org/pypy/extradoc/src/tip/talk/vmil2012}}
All
benchmark data was collected on a MacBook Pro 64 bit running Max OS 10.8 with
the loop unrolling optimization disabled.\footnote{Since loop unrolling
duplicates the body of loops it would no longer be possible to meaningfully
compare the number of operations before and after optimization. Loop unrolling
is most effective for numeric kernels, so the benchmarks presented here are not
affected much by its absence.}

@@

The overhead that is incurred by the JIT to manage the \texttt{resume data},
the \texttt{low-level resume data} as well as the generated machine code is
shown in Figure~\ref{fig:backend_data}. It shows the total memory consumption
of the code and of the data generated by the low-level component and an
approximation of the size of the \texttt{resume data} structures for the
different benchmarks mentioned above. The machine code taken into account is
composed of the compiled operations, the trampolines generated for the guards
and a set of support functions that are generated when the JIT starts and which
are shared by all compiled traces. The size of the \texttt{low-level resume
data} is the size of the compressed mapping from registers and stack to
IR-level variables and finally the size of the \texttt{resume data} is an
approximation of the size of the compressed high-level resume data as described
in Section~\ref{sec:Resume Data}.\footnote{
The size of the resume data is not measured at runtime, but reconstructed from
log files.}
